---
title: "Why Coordinates Matter - Demonstration"
author: "Nick Hoffman"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    highlight: pygment
    keep_md: no
    toc: true
    number_sections: true
    toc_depth: 1
    toc_float: true
    theme: journal
editor_options:
    chunk_output_type: inline
---

<style type="text/css">
h2, h3, h4, h5, h6 {
  counter-reset: section;
}
p {
  font-size:18px;
}

ul {
  font-size:18px;
}

li {
  font-size:18px;
}
table {
   padding: 0;border-collapse: collapse;
   layout: fixed;
   width: 90%; }
table tr {
   border-top: 1px solid #cccccc;
   background-color: white;
   margin: 0;
   padding: 0; }
table tr:nth-child(2n) {
   background-color: #f8f8f8; }
table tr th {
   font-weight: bold;
   border: 1px solid #cccccc;
   margin: 0;
   padding: 6px 13px; }
table tr td {
   border: 1px solid #cccccc;
   margin: 0;
   padding: 6px 13px; }
table tr th :first-child, table tr td :first-child {
   margin-top: 0; }
table tr th :last-child, table tr td :last-child {
   margin-bottom: 0; }
.html-widget {
    margin: auto;
}
</style>

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This page will demonstrate one situation where the difference between site and core location coordinates matters when using paleodata from the Neotoma Paleoecology Database.

```{r packages, warning=FALSE, message=FALSE,echo=FALSE}

library(neotoma2)
library(tidyverse)
library(DT)
library(geojsonsf)
library(sf)
library(leaflet)
library(jsonlite)
library(httr)
library(ggplot2)
library(analogue)
#library(geodata)
#library(elevatr)
library(raster)
library(tmap)

```

# Our sites

Our sites and cores are mapped out below.

```{r get-data, echo=FALSE, warning=FALSE}

tanganyikaSites = get_sites(sitename = "%Tanganyika%")

```


```{r api-sites, echo=FALSE, warning=FALSE}


siteidString = paste0(as.data.frame(tanganyikaSites)$siteid,collapse=",")

apiCall = paste0('https://api.neotomadb.org/v2.0/data/sites/',siteidString,'/datasets')


response = GET(apiCall)

siteMetadata = content(response)$data

```



```{r metadata-to-table, warning=FALSE,echo=FALSE}

idx = 0

for (i in seq(length(siteMetadata))) {
  for (j in seq(length(siteMetadata[[i]]$site$datasets))) {
    idx = idx + 1
  }
}


siteMetadata_mat = matrix(nrow=idx,ncol=10)

idx2 = 0
for (i in seq(length(siteMetadata))) {
  for (j in seq(length(siteMetadata[[i]]$site$datasets))) {
    idx2 = idx2 + 1
    for (k in seq(10)) {
      if (!is.null(siteMetadata[[i]]$site[[k]])) {
        siteMetadata_mat[[idx2, k]] = siteMetadata[[i]]$site[[k]]
      }
    }
  }
}


siteMetadata_df = as.data.frame(siteMetadata_mat)

names(siteMetadata_df) = c("siteid","sitename","sitedescription","sitenotes","geography","altitude","collectionunitid","collectionunit","handle","unittype")


siteMetadata_df = distinct(siteMetadata_df)



```

```{r geo-revisited, warning=FALSE, echo=FALSE}



siteGeo_sf = geojson_sf(siteMetadata_df$geography)

siteMetadata_sf = cbind(siteGeo_sf,siteMetadata_df)

pointSites = siteMetadata_sf[st_geometry_type(siteMetadata_sf) == "POINT",] %>% distinct(siteid, .keep_all = TRUE)
polySites = siteMetadata_sf[st_geometry_type(siteMetadata_sf) == "POLYGON",] %>% distinct(siteid, .keep_all = TRUE)


```



```{r collectionunits, warning=FALSE, echo=FALSE}


text="collectionunits"
collunits = content(GET(paste0('https://api.neotomadb.org/v2.0/data/dbtables/',text,'?count=false&limit=99999&offset=0')))$data



collunit_mat = matrix(nrow=length(collunits),ncol=20)

for (i in seq(length(collunits))) {
  for (j in seq(20)) {
    if (!is.null(collunits[[i]][[j]])) {
      collunit_mat[[i,j]] = collunits[[i]][[j]]
    }
  }
}

collunit_df = collunit_mat %>% as.data.frame()

names(collunit_df) = c("collectionunitid","handle","siteid","colltypeid","depenvtid","collunitname","colldate","colldevice","gpslatitude","gpslongitude","gpsaltitude","gpserror","waterdepth","substrateid","slopeaspect","slopeangle","location","notes","recdaterecreated","recdatemodified")


filtered_colls = collunit_df %>% dplyr::filter(collectionunitid %in% siteMetadata_df$collectionunitid) 
filteredColl_sf = collunit_df %>% dplyr::filter(collectionunitid %in% siteMetadata_df$collectionunitid) %>% st_as_sf(coords=c("gpslongitude","gpslatitude"), crs="WGS84")



pal <- colorFactor(palette = "Set1", domain = siteMetadata_df$siteid)


polySites$arranger =  cbind( 1/as.numeric(st_area(polySites)))
polySites = polySites %>% arrange(arranger)

leaflet() %>%
  addTiles() %>%
  addPolygons(data = polySites, 
              color = "red", 
              weight = 2, 
              fillColor = ~pal(siteid), 
              fillOpacity = 0.55, 
              popup = ~siteid)  %>%
  addCircleMarkers(data = pointSites, 
                   radius = 5, 
                   color = "black", 
                   fillColor = ~pal(siteid), 
                   fillOpacity = 0.7, 
                   stroke = TRUE, 
                   weight = 1,
                   popup = ~siteid)   %>% 
  addCircleMarkers(data = filteredColl_sf, 
                   radius = 3, 
                   color = "white", 
                   fillColor = ~pal(siteid), 
                   fillOpacity = 1, 
                   stroke = TRUE,
                   weight = 1,
                   popup = ~collectionunitid)

```

# Depths

```{r depths}

samps = samples(get_downloads(get_datasets(tanganyikaSites)))

pollen = samps %>% dplyr::filter(datasettype=="pollen" & element == "pollen")

site_sf = as.data.frame(tanganyikaSites)[c(4,3,2,1)] %>% st_as_sf(coords=c("long","lat"), crs="WGS84")

bathy = raster("tanganyika_bathy.tif")


coll_depth = raster::extract(bathy,st_coordinates(filteredColl_sf))
site_depth = raster::extract(bathy,st_coordinates(site_sf))


print(coll_depth)
print(site_depth)


fromneo = pollen %>% group_by(siteid,collunitid) %>% distinct(elev)
print(fromneo)
```

# Sedge abundance


```{r exercise, message=FALSE,results='hide'}



toptaxa = pollen %>% group_by(variablename) %>% count() %>% arrange(desc(n)) %>% head(n=6)

count_coll = pollen %>% dplyr::filter(variablename %in% toptaxa$variablename) %>% dplyr::select(collunitid,age,variablename,value) %>%
  group_by(collunitid,age) %>% 
  dplyr::mutate(pollcount = sum(value, na.rm=TRUE)) %>%
  ungroup() %>%
  group_by(collunitid,age,variablename) %>%
  mutate(sum_poll = sum(value,na.rm=TRUE)) %>%
  ungroup() %>%
  group_by(variablename) %>%
  mutate(prop = sum_poll / pollcount) %>%
  dplyr::select(collunitid, age, variablename, prop) %>% 
  mutate(prop = as.numeric(prop)) %>%
  drop_na(age) %>% 
  distinct()

counts_coll = tidyr::pivot_wider(count_coll,
                             id_cols = c(age,collunitid),
                             names_from = variablename,
                             values_from = prop,
                             values_fill = 0)


counts_coll = counts_coll %>% left_join(filtered_colls, by = join_by(collunitid == collectionunitid))

count_site = pollen %>% dplyr::filter(variablename %in% toptaxa$variablename) %>% dplyr::select(siteid,age,variablename,value) %>% 
  group_by(siteid,age) %>% 
  dplyr::mutate(pollcount = sum(value, na.rm=TRUE)) %>%
  ungroup() %>%
  group_by(siteid,age,variablename) %>%
  mutate(sum_poll = sum(value,na.rm=TRUE)) %>%
  ungroup() %>%
  group_by(variablename) %>%
  mutate(prop = sum_poll / pollcount) %>%
  dplyr::select(siteid, age, variablename, prop) %>% 
  mutate(prop = as.numeric(prop)) %>%
  drop_na(age) %>% 
  distinct()

counts_site = tidyr::pivot_wider(count_site,
                             id_cols = c(age,siteid),
                             names_from = variablename,
                             values_from = prop,
                             values_fill = 0)


counts_site = counts_site %>% left_join(as.data.frame(tanganyikaSites), by=join_by(siteid==siteid))


counts_coll = counts_coll %>% dplyr::mutate(gpslatitude = round(as.numeric(gpslatitude),6))
counts_coll = counts_coll %>% dplyr::mutate(gpslongitude = round(as.numeric(gpslongitude),6))


#clim2 <- geodata::worldclim_global(var = 'bio', res = 0.5, path=tempdir())

#lc_tree = landcover(var='trees',path=tempdir())
#lc_grass =  landcover(var='grassland',path=tempdir())




lats = c(-3,-3,-9,-9) %>% as.data.frame()
lons = c(28,32,28,32) %>% as.data.frame()

coordinates = lats %>% cbind(lons) 
names(coordinates) = c("lat","lon")

coordinates = coordinates %>% st_as_sf(coords=c("lon","lat"), crs="WGS84") %>% mutate(dummyID = 1) %>% group_by(dummyID) %>% dplyr::summarize() %>% st_cast("POLYGON")

#lc_tree_raster <- raster("lc_eraster.tif")

#lc_grass_raster <- raster("lc_grass_raster.tif")

clim30_subset=stack("clim30_subset.tif")

#coll_map = raster::extract(clim30_12,st_coordinates(filteredColl_sf))
#site_map = raster::extract(clim30_12,as.data.frame(tanganyikaSites)[c(4,3)])



coll_bioclim = raster::extract(clim30_subset,st_coordinates(filteredColl_sf)) %>% cbind(filteredColl_sf)
site_bioclim = raster::extract(clim30_subset,as.data.frame(tanganyikaSites)[c(4,3)]) %>% cbind(as.data.frame(tanganyikaSites))



counts_site_young = counts_site %>% group_by(siteid) %>% dplyr::filter(age == min(age)) %>% left_join(site_bioclim)
counts_coll_young =  counts_coll %>% group_by(collunitid) %>% dplyr::filter(age == min(age)) %>% left_join(coll_bioclim)


names(counts_coll_young) = c("age","collunitid","Olea capensis","Brachystegia","Alchornea","Podocarpus","Cyperaceae","Poaceae","handle","siteid","colltypeid","depenvtid","collunitname","colldate","colldevice","lat","lon","altitude","gpserror","waterdepth","substrateid","slopeaspect","slopeangle","location","notes","datecreated","datemodified","bio1","bio2","bio3","bio4","bio5","bio6","bio7","bio8","bio9","bio10","bio11","bio12","bio13","bio14","bio15","bio16","bio17","bio18","bio19","collectionunitid","geometry")

names(counts_site_young) = c("age","siteid","Olea capensis","Brachystegia","Alchornea","Podocarpus","Cyperaceae","Poaceae","sitename", "lat","lon","area","notes","description","elev","bio1","bio2","bio3","bio4","bio5","bio6","bio7","bio8","bio9","bio10","bio11","bio12","bio13","bio14","bio15","bio16","bio17","bio18","bio19")

names(clim30_subset) = c("bio1","bio2","bio3","bio4","bio5","bio6","bio7","bio8","bio9","bio10","bio11","bio12","bio13","bio14","bio15","bio16","bio17","bio18","bio19")

## 11 vars is fast
## 15 takes 4 or 5 minutes - alg didn't converge


coll_mod = glm(Cyperaceae ~ (bio1 + bio2 + bio3 + bio4 + bio5 + bio6 + bio7 + bio8 + bio9 + bio10+ bio11+ bio12+ bio13+ bio14 + bio15 + bio16 + bio17 + bio18 + bio19)^2, data=counts_coll_young)

coll_step = step(coll_mod)


site_mod = glm(Cyperaceae ~ (bio1 + bio2 + bio3 + bio4 + bio5 + bio6 + bio7+ bio8+ bio9+ bio10+ bio11+ bio12+ bio13+ bio14 + bio15 + bio16 + bio17 + bio18 + bio19)^2, data=counts_site_young)

site_step = step(site_mod)


coll_pred = predict(clim30_subset,model=coll_step)


site_pred = predict(clim30_subset,model=site_step)

difference = coll_pred - site_pred



min_val <- minValue(difference)
max_val <- maxValue(difference)
breaks <- seq((0-max(max_val,(0-min_val))),max(max_val,(0-min_val)),length.out=100)

color_palette <- colorRampPalette(c("blue", "white", "red"))(99)

plot(coll_pred)
plot(site_pred)
plot(difference, col=color_palette,breaks=breaks)


## all equal 0
##coll_lc = raster::extract(lc_raster,st_coordinates(filteredColl_sf))
##site_lc = raster::extract(lc_raster,as.data.frame(tanganyikaSites)[c(4,3)])
## all equal 0
#coll_grass = raster::extract(lc_grass_raster,st_coordinates(filteredColl_sf))
#site_grass = raster::extract(lc_grass_raster,as.data.frame(tanganyikaSites)[c(4,3)])

#depth_raster = get_aws_terrain(coordinates,prj=st_crs(coordinates),z=5)

depth_raster = raster("depth_raster.tif")
site_depth = raster::extract(depth_raster,as.data.frame(tanganyikaSites)[c(4,3)])



site_sf = counts_site %>% st_as_sf(coords=c("long","lat"),crs="WGS84") %>% dplyr::select(siteid) %>% distinct()


buffer_radius <- 25000  # Radius in meters
coll_circle <- st_buffer(filteredColl_sf, dist = buffer_radius)
site_circle <- st_buffer(site_sf, dist = buffer_radius)


extent_region <- extent(coordinates) # Replace with your area coordinates

grass_lc_subset = raster("grass_lc_subset.tif")


tm_shape(grass_lc_subset) + tm_raster(style="cont") + 
    tm_shape(filteredColl_sf) + tm_dots(col="red",size=0.5,alpha=0.5) +
    tm_shape(site_sf) + tm_dots(col="black",size=0.5,alpha=0.5) +
    tm_shape(coll_circle) + tm_borders(col="red",alpha=0.5) +
    tm_shape(site_circle) + tm_borders(col="black",alpha=0.5)



# Function to crop, mask, and calculate mean for each polygon
calculate_mean_grass <- function(poly) {
  cropped <- raster::crop(grass_lc_subset, as_Spatial(poly))          # Crop to bounding box
  masked <- raster::mask(cropped, as_Spatial(poly))               # Mask to polygon shape
  mean_val <- cellStats(masked, stat = 'mean', na.rm = TRUE)  # Calculate mean
  return(mean_val)
}




mean_coll_grass <- lapply(1:nrow(coll_circle), function(i) calculate_mean_grass(coll_circle[i, ]))
mean_coll_grass <- unlist(mean_coll_grass)  # Convert to numeric vector if needed


mean_site_grass <- lapply(1:nrow(site_circle), function(i) calculate_mean_grass(site_circle[i, ]))
mean_site_grass <- unlist(mean_site_grass)  # Convert to numeric vector if needed



coll_circle = coll_circle %>% cbind(mean_coll_grass)

site_circle = site_circle %>% cbind(mean_site_grass)



counts_coll = counts_coll %>% dplyr::left_join(coll_circle, by=join_by(collunitid == collectionunitid)) 

counts_site = counts_site %>% dplyr::left_join(site_circle, by=join_by(siteid == siteid)) 


counts_coll_young = counts_coll %>% dplyr::filter(age <100) %>% pivot_longer(cols=c("Olea capensis","Brachystegia","Alchornea","Podocarpus","Cyperaceae","Poaceae"),names_to="taxon",values_to="percent")
counts_site_young = counts_site  %>% dplyr::filter(age <100) %>% pivot_longer(cols=c("Olea capensis","Brachystegia","Alchornea","Podocarpus","Cyperaceae","Poaceae"),names_to="taxon",values_to="percent")

ggplot() +
  geom_point(data = counts_coll_young, mapping=aes(x=mean_coll_grass,y=percent),col="red") +
 # geom_smooth(data = counts_coll_young, mapping=aes(x=mean_coll_grass,y=percent),col="red") +
  geom_point(data = counts_site_young, mapping=aes(x=mean_site_grass,y=percent),col="blue") +
  # geom_smooth(data = counts_site_young, mapping=aes(x=mean_site_grass,y=percent),col="blue") +
  facet_wrap(~taxon,scales="free")


```

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>