---
title: "Taxon Harmonization"
author: "Nick Hoffman"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    highlight: pygment
    keep_md: no
    toc: true
    number_sections: true
    toc_depth: 1
    toc_float: true
    theme: journal
editor_options:
    chunk_output_type: inline
---

<style type="text/css">
h2, h3, h4, h5, h6 {
  counter-reset: section;
}
p {
  font-size:18px;
}

ul {
  font-size:18px;
}

li {
  font-size:18px;
}
table {
   padding: 0;border-collapse: collapse;
   layout: fixed;
   width: 90%; }
table tr {
   border-top: 1px solid #cccccc;
   background-color: white;
   margin: 0;
   padding: 0; }
table tr:nth-child(2n) {
   background-color: #f8f8f8; }
table tr th {
   font-weight: bold;
   border: 1px solid #cccccc;
   margin: 0;
   padding: 6px 13px; }
table tr td {
   border: 1px solid #cccccc;
   margin: 0;
   padding: 6px 13px; }
table tr th :first-child, table tr td :first-child {
   margin-top: 0; }
table tr th :last-child, table tr td :last-child {
   margin-bottom: 0; }
.html-widget {
    margin: auto;
}
</style>

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages,message=FALSE,warning=FALSE}

library(neotoma2)
library(tidyverse)
library(sf)
library(geojsonsf)
library(httr)
library(jsonlite)
library(stringr)
library(ggplot2)
```

```{r get-data}

lats = c(-16,-16,-36,-36) %>% as.data.frame()
lons = c(10,41,10,41) %>% as.data.frame()

coordinates = lats %>% cbind(lons) 
names(coordinates) = c("lat","lon")

coordinates = coordinates %>% st_as_sf(coords=c("lon","lat"), crs="WGS84") %>% mutate(dummyID = 1) %>% group_by(dummyID) %>% dplyr::summarize() %>% st_cast("POLYGON")

coord_json = sf_geojson(coordinates)


altmin = 000 #m

sites = content(GET(paste0("https://api.neotomadb.org/v2.0/data/sites?altmin=",altmin,"&loc=",coord_json,"&limit=999&offset=0")))$data


idx = 0
for (i in seq(length(sites))) {
  for (j in seq(length(sites[[i]]$collectionunits))) {
    for (k in seq(length(sites[[i]]$collectionunits[[j]]$datasets))) {
    idx = idx + 1
    }
    }
}


sites_mat = matrix(nrow=idx,ncol=11)

idx2 = 0
for (i in seq(length(sites))) {
  for (j in seq(length(sites[[i]]$collectionunits))) {
    for (k in seq(length(sites[[i]]$collectionunits[[j]]$datasets))) {
    idx2 = idx2 + 1
    for (m in seq(5)) {
      if (!is.null(sites[[i]][[m]])) {
        sites_mat[[idx2, m]] = sites[[i]][[m]]
      }
    }
    
     if (!is.null(sites[[i]]$collectionunits[[j]]$handle)) {
        sites_mat[[idx2,6]] = sites[[i]]$collectionunits[[j]]$handle
     }
       if (!is.null(sites[[i]]$collectionunits[[j]]$collectionunit)) {
        sites_mat[[idx2,7]] = sites[[i]]$collectionunits[[j]]$collectionunit
       }
       if (!is.null(sites[[i]]$collectionunits[[j]]$collectionunitid)) {
        sites_mat[[idx2,8]] = sites[[i]]$collectionunits[[j]]$collectionunitid
       }
       if (!is.null(sites[[i]]$collectionunits[[j]]$collectionunittype)) {
        sites_mat[[idx2,9]] = sites[[i]]$collectionunits[[j]]$collectionunittype
       }
       if (!is.null(sites[[i]]$collectionunits[[j]]$dataset[[k]]$datasetid)) {
        sites_mat[[idx2,10]] = sites[[i]]$collectionunits[[j]]$dataset[[k]]$datasetid
       }
       if (!is.null(sites[[i]]$collectionunits[[j]]$dataset[[k]]$datasettype)) {
        sites_mat[[idx2,11]] = sites[[i]]$collectionunits[[j]]$dataset[[k]]$datasettype
       }
    }
  }
}

sites_df = as.data.frame(sites_mat)

names(sites_df) = c("siteid","sitename","sitedescription","geography","altitude","handle","collectionunit","collectionunitid","collectionunittype","datasetid","datasettype")

siteids = sites_df %>% distinct(siteid) %>% dplyr::mutate()


sites_neo = get_sites(as.numeric(siteids$siteid),all_data=TRUE)

data = samples(get_downloads(get_datasets(sites_neo,all_data=TRUE),all_data=TRUE))

pollendata = data %>% dplyr::filter(datasettype=="pollen")

pollendata %>% group_by(variablename) %>% count() %>% arrange(desc(n))
  

apd_harmTable = read.csv("APD_dictionnary_export.csv",row.names=NULL,sep=";")

pollendata = pollendata %>% left_join(apd_harmTable, by=join_by(variablename == Taxon..original.name.))

pollendata = pollendata %>% mutate(Taxon..revised.nomenclature. = case_when(
  str_detect(variablename,regex('poaceae', ignore_case = T)) ~ "Poaceae undiff.", 
  TRUE ~ Taxon..revised.nomenclature.)) %>% mutate(Family.name = case_when(
  str_detect(variablename,regex('poaceae', ignore_case = T)) ~ "POACEAE", 
  TRUE ~ Family.name))


pollen_alt = pollendata %>% mutate(elevbins = round(elev,-2)) %>% drop_na(age) %>% mutate(agebins = round(age/5000)*5000)

pollen_alt = pollen_alt %>% group_by(elevbins,Family.name,agebins) %>% summarize(polcount = sum(value)) 

pol_total = pollen_alt %>% ungroup() %>% group_by(elevbins,agebins) %>% summarize(total = sum(polcount))

pollen_alt = pollen_alt %>% left_join(pol_total) %>% mutate(prop = polcount/total*100) %>% drop_na()

wide_alt = pollen_alt %>% pivot_wider(id_cols=elevbins,names_from=Family.name,values_from=prop)

pollen_alt_young = pollen_alt %>% dplyr::filter(agebins < 29000) 
toptaxa = pollen_alt_young %>% group_by(Family.name) %>% dplyr::summarize(propsum = sum(prop)) %>% arrange(desc(propsum)) %>% head(n=10)

pollen_alt_young = pollen_alt_young %>% dplyr::filter(Family.name %in% toptaxa$Family.name)

ggplot(pollen_alt_young) +
  geom_line(mapping=aes(x=elevbins,y=log(prop),group=agebins,color=as.factor(agebins)),lwd=2) + 
  facet_wrap(~Family.name) 


```